\section{Introduction}
Concurrent programming is an important part of modern software because many programs need to use more than one thread at the same time. Threads may run in parallel, share data, communicate through memory, and update the same objects. This can improve performance, but it also makes the program harder to reason about. If shared data is not handled correctly, the program may produce wrong results even if each thread looks correct on its own.

One common way to make concurrent programs easier to use is through concurrent data structures. These are data structures that are designed to be accessed by multiple threads without breaking their internal state. Instead of forcing every user of the structure to write synchronization code manually, the synchronization is built into the data structure itself. This is useful because structures such as queues, stacks, maps, and lists are often shared between worker threads.

Queues are especially useful in concurrent programs. They are often used to pass work from producer threads to consumer threads, to store tasks, or to connect different parts of a system. A concurrent queue must preserve the normal FIFO behavior while also allowing multiple threads to insert and remove elements safely. This makes queues a good case study for understanding synchronization, lock-free algorithms, and the tradeoffs between correctness and performance.

Java already provides concurrent queue implementations in the JDK. One important example is \texttt{ConcurrentLinkedQueue}, which is a non-blocking FIFO queue based on ideas from the Michael and Scott queue algorithm. Because this queue is part of the standard library, it is a useful baseline for comparison. It gives a real implementation that is already used in Java programs, so custom implementations can be compared against something practical instead of only against a simple theoretical example.

The goal of this project is to implement and study lock-free queue approaches in Java, while also understanding concepts such as \texttt{volatile}, \texttt{VarHandle}s, CAS operations, visibility, atomicity, and the Java Memory Model. The first custom implementation is inspired by the Michael-Scott queue and uses linked nodes, \texttt{volatile} references, \texttt{VarHandle}s, and compare-and-set operations. The second implementation is inspired by the Linked Concurrent Ring Queue idea. It is not a full implementation of the original LCRQ algorithm, but it uses similar ideas such as linked segments, array-based storage, and atomic operations inside each segment.

Both custom queues were implemented with the same main API as the JDK queue, so they can be tested in the same way. The implementations were added inside JDK 21 and are executed through that version for the main experiments. The standalone source code is also kept in the project so that the queue implementations can be inspected or run without building the modified JDK. This allows the custom queues to be studied directly together with the existing JDK queue implementations.

The project also includes correctness testing and performance evaluation. Correctness is checked by comparing the behavior of the custom queues with the official JDK queue and by testing concurrent behavior. Evaluation is then used to study how the different implementations behave under different thread counts and workloads. The main purpose is to understand the design choices behind these queues and to see how the custom implementations compare with the existing JDK implementation.
